Generating Random Logic Programs Using Constraint Programming

### 2020-08-14 conference_talk/talk.tex
More slides
Move things around

\documentclass{beamer}
\usepackage{tikz}
\usepackage[style=authoryear]{biblatex}
\usepackage{empheq}

\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes}
\tikzstyle{every picture}+=[remember picture]

@@

@@

\begin{frame}{Inference Algorithms and Knowledge Compilation Maps}
  \begin{description}
  \item[NNF] negation normal form
  \item[d-DNNF] deterministic decomposable negation normal form
  \item[BDD] binary decision diagrams
  \item[SDD] sentential decision diagrams
  \item[$k$-Best] only use the $k$ most probable proofs
  \end{description}
\end{frame}

\begin{frame}{Example Diagrams for $C \land (A \lor \neg B)$}
  \vspace{-1em}
  % \begin{tikzpicture}
  %   \node[draw,circle] {$\land$}
  %   child {node[draw,rectangle] {$C$}}
  %   child {node[draw,circle] {$\lor$}
  %     child {node[draw,rectangle] {$A$}}
  %     child {node[draw,rectangle] {$\neg B$}
  %     }
  %   };
  % \end{tikzpicture}
  \begin{columns}[b]
    \begin{column}{0.33\textwidth}
      \begin{figure}
      \centering
        \begin{tikzpicture}
          \node[draw,circle] (A) {$A$}
          child {edge from parent[draw=none]}
          child {node[draw,circle] (B) {$B$} edge from parent[dashed]
            child {node[draw,circle,solid] (C) {$C$} edge from parent[dashed]
              child {node[draw,rectangle,solid] {$1$} edge from parent[solid]}
              child {node[draw,rectangle,solid] (0) {$0$}}
            }
            child {edge from parent[draw=none]}
          };
          \draw (A) -- (C);
          \draw (B) -- (0);
        \end{tikzpicture}
        \caption{BDD}
      \end{figure}
    \end{column}
    \begin{column}{0.34\textwidth}
      \begin{figure}
        \centering
        \begin{tikzpicture}
          \node[draw,circle] {$\land$}
          child {node[draw,circle] {$\lor$}
            child {node[draw,circle] (parent) {$\land$}
              child {node[draw,rectangle] {$B$}}
              child {edge from parent[draw=none]}
            }
            child {node[draw,circle] {$\land$}
              child {node[draw,circle] {$\lor$}
                child {node[draw,rectangle] (A) {$A$}}
                child {node[draw,rectangle] {$\neg A$}}
              }
              child {node[draw,rectangle] {$\neg B$}}
            }
          }
          child {node[draw,rectangle] {$C$}};
          \draw (parent) -- (A);
        \end{tikzpicture}
        \caption{d-DNNF}
      \end{figure}
    \end{column}
    \begin{column}{0.33\textwidth}
      \begin{figure}
        \centering
        \begin{tikzpicture}
          \tikzset{
            mysplit/.style={
              draw,
              rectangle,
              rectangle split,
              rectangle split horizontal,
              rectangle split parts=2
            }
          }
          \node[draw,circle] {$1$}
          child {node[mysplit] (bullet) {
              \nodepart{one} $\neg A$
              \nodepart{two}
            }
            child {node[draw,circle] (3) {$3$} edge from parent[draw=none]
              child {node[mysplit] {
                  \nodepart{one} $\neg B$
                  \nodepart{two} $C$
                }
              }
              child {node[mysplit] {
                  \nodepart{one} $B$
                  \nodepart{two} $0$
                }
              }
            }
          }
          child {node[mysplit] {
              \nodepart{one} $A$
              \nodepart{two} $C$
            }};
          \draw[*-] let \p1 = (bullet.two), \p2 = (bullet.center) in ({\x1 + 2.5},{\y2 + 2}) -- (3);
        \end{tikzpicture}
        \caption{SDD}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

@@

@@

\begin{frame}{Clauses As Trees}
  \[
    \neg \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})
    \lor (\textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}})
    \land \textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{X}}))
  \]
  \vfill
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \pause
      \centering
      \begin{tikzpicture}[level 1/.style={sibling distance=30mm},level 2/.style={sibling distance=15mm}]
        \node[draw,circle] {$\lor$}
        child {node[draw,circle] {$\neg$}
          child {node[draw,rectangle] {$\textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})$}}
        }
        child {node[draw,circle] {$\land$}
          child {node[draw,rectangle] {$\textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}})$}}
          child {node[draw,rectangle] {$\textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{X}})$}}
        };
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.5\textwidth}
      \pause
      \centering
      \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        0 & 0 & 0 & 1 & 2 & 2 \\
        \hline
        $\lor$ & $\neg$ & $\land$ & $\textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})$ & $\textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}})$ & $\textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{X}})$ \\
        \hline
      \end{tabular}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{Independence: $\textcolor{predicate}{\mathrm{influences}}
    \independent \textcolor{predicate}{\mathrm{stress}}$}
  \centering
  \begin{tikzpicture}
    \node[draw,ellipse] (friend) {\textcolor{predicate}{friend}};
    \node[draw,ellipse,right=of friend] (influences) {\textcolor{predicate}{influences}};
    \node[draw,ellipse,below=of influences] (stress) {\textcolor{predicate}{stress}};
    \node[draw,ellipse,below=of stress] (spont) {\textcolor{predicate}{cancer\_spont}};
    \node[draw,ellipse,below=of spont] (smoke) {\textcolor{predicate}{cancer\_smoke}};
    \node[draw,ellipse,left=of spont] (person) {\textcolor{predicate}{person}};
    \node[draw,ellipse,right=of spont] (cancer) {\textcolor{predicate}{cancer}};
    \node[draw,ellipse,above=of cancer] (smokes) {\textcolor{predicate}{smokes}};
    \draw[-{Latex[length=2mm]}] (friend) -- (influences);
    \draw[-{Latex[length=2mm]}] (influences) -- (smokes);
    \draw[-{Latex[length=2mm]}] (smokes) -- (cancer);
    \draw[-{Latex[length=2mm]}] (stress) -- (smokes);
    \draw[-{Latex[length=2mm]}] (spont) -- (cancer);
    \draw[-{Latex[length=2mm]}] (smoke) -- (cancer);
    \draw[-{Latex[length=2mm]}] (person) -- (stress);
    \draw[-{Latex[length=2mm]}] (person) -- (spont);
    \draw[-{Latex[length=2mm]}] (person) -- (smoke);
    \draw<2->[-{Latex[length=2mm]}] (influences) edge[draw=red,bend left=15] (stress);
    \draw<2->[-{Latex[length=2mm]}] (stress) edge[draw=red,bend left=15] (influences);
    \draw<2->[-{Latex[length=2mm]}] (stress) edge[draw=red,bend left=15] (friend);
    \draw<2->[-{Latex[length=2mm]}] (friend) edge[draw=red,bend left=15] (stress);
  \end{tikzpicture}
\end{frame}

\begin{frame}{Stratification and Negative Cycles}
  \centering
  \begin{tikzpicture}
    \node[draw,ellipse] (friend) {\textcolor{predicate}{friend}};
    \node[draw,ellipse,right=of friend] (influences) {\textcolor{predicate}{influences}};
    \node[draw,ellipse,below=of influences] (stress) {\textcolor{predicate}{stress}};
    \node[draw,ellipse,below=of stress] (spont) {\textcolor{predicate}{cancer\_spont}};
    \node[draw,ellipse,below=of spont] (smoke) {\textcolor{predicate}{cancer\_smoke}};
    \node[draw,ellipse,left=of spont] (person) {\textcolor{predicate}{person}};
    \node[draw,ellipse,right=of spont] (cancer) {\textcolor{predicate}{cancer}};
    \node[draw,ellipse,above=of cancer] (smokes) {\textcolor{predicate}{smokes}};
    \draw[-{Latex[length=2mm]}] (friend) -- (influences);
    \draw[-{Latex[length=2mm]}] (influences) -- (smokes);
    \draw[-{Latex[length=2mm]}] (smokes) -- (cancer);
    \draw[-{Latex[length=2mm]}] (stress) -- (smokes);
    \draw[-{Latex[length=2mm]}] (spont) -- (cancer);
    \draw[-{Latex[length=2mm]}] (smoke) -- (cancer);
    \draw[-{Latex[length=2mm]}] (person) -- (stress);
    \draw[-{Latex[length=2mm]}] (person) -- (spont);
    \draw[-{Latex[length=2mm]}] (person) -- (smoke);
    \draw<2->[-{Latex[length=2mm]},draw=red] (smokes) -- (friend);
  \end{tikzpicture}
  \[
    \textcolor{probability}{0.1} \prob
    \textcolor{predicate}{\mathtt{friend}}(\textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}}) \ifff
    \negg\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}).
  \]
\end{frame}

@@

@@

% TODO: finish the slide
\begin{frame}{Summary}
  \begin{itemize}
  \item foo
  \item bar
  \item baz
  \end{itemize}
  \begin{block}{The implementation of the model is available at}
    \url{https://github.com/dilkas/random-logic-programs}
  \end{block}
\end{frame}

>>>>

\documentclass{beamer}
\usepackage{tikz}
\usepackage[style=authoryear,backend=bibtex]{biblatex}
\usepackage{empheq}

\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes}
\tikzstyle{every picture}+=[remember picture]
\tikzset{onslide/.code args={<#1>#2}{%
  \only<#1>{\pgfkeysalso{#2}}
}}

@@

\section{Introduction}

@@

@@

\begin{frame}
  \frametitle{Outline}
  \tableofcontents
\end{frame}

\section{The Constraint Model}

@@

\begin{frame}{Formulas As Trees}
  \[
    \neg \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})
    \lor (\textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}})
    \land \textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{X}}))
  \]
  \vfill
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \pause
      \centering
      \begin{tikzpicture}[level 1/.style={sibling distance=30mm},level 2/.style={sibling distance=15mm}]
        \node[draw,circle] {$\lor$}
        child {node[draw,circle] {$\neg$}
          child {node[draw,rectangle] {$\textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})$}}
        }
        child {node[draw,circle] {$\land$}
          child {node[draw,rectangle] {$\textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}})$}}
          child {node[draw,rectangle] {$\textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{X}})$}}
        };
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.5\textwidth}
      \pause
      \centering
      \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        0 & 0 & 0 & 1 & 2 & 2 \\
        \hline
        $\lor$ & $\neg$ & $\land$ & $\textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})$ & $\textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}})$ & $\textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{X}})$ \\
        \hline
      \end{tabular}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{Independence: $\textcolor{predicate}{\mathrm{friend}} \independent
    \textcolor{predicate}{\mathrm{stress}}$}
  \centering
  \begin{tikzpicture}
    \node[draw,ellipse,onslide=<2->{fill=constant!25}] (friend) {\textcolor{predicate}{friend}};
    \node[draw,ellipse,right=of friend] (influences) {\textcolor{predicate}{influences}};
    \node[draw,ellipse,onslide=<2->{fill=variable!25},below=of influences] (stress) {\textcolor{predicate}{stress}};
    \node[draw,ellipse,below=of stress] (spont) {\textcolor{predicate}{cancer\_spont}};
    \node[draw,ellipse,below=of spont] (smoke) {\textcolor{predicate}{cancer\_smoke}};
    \node[draw,ellipse,onslide=<2->{fill=variable!25},left=of spont] (person) {\textcolor{predicate}{person}};
    \node[draw,ellipse,right=of spont] (cancer) {\textcolor{predicate}{cancer}};
    \node[draw,ellipse,above=of cancer] (smokes) {\textcolor{predicate}{smokes}};
    \draw[-{Latex[length=2mm]}] (friend) -- (influences);
    \draw[-{Latex[length=2mm]}] (influences) -- (smokes);
    \draw[-{Latex[length=2mm]}] (smokes) -- (cancer);
    \draw[-{Latex[length=2mm]}] (stress) -- (smokes);
    \draw[-{Latex[length=2mm]}] (spont) -- (cancer);
    \draw[-{Latex[length=2mm]}] (smoke) -- (cancer);
    \draw[-{Latex[length=2mm]}] (person) -- (stress);
    \draw[-{Latex[length=2mm]}] (person) -- (spont);
    \draw[-{Latex[length=2mm]}] (person) -- (smoke);

    \draw<3->[-{Latex[length=2mm]}] (friend) edge[draw=red,bend left=15] (person);
    \draw<3->[-{Latex[length=2mm]}] (person) edge[draw=red,bend left=15] (friend);

    \draw<3->[-{Latex[length=2mm]}] (stress) edge[draw=red,bend left=15] (friend);
    \draw<3->[-{Latex[length=2mm]}] (friend) edge[draw=red,bend left=15] (stress);
  \end{tikzpicture}
\end{frame}

\begin{frame}{Stratification and Negative Cycles}
  \[
    \textcolor{probability}{0.1} \prob
    \textcolor{predicate}{\mathtt{friend}}(\textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}}) \ifff
    \negg\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}).
  \]
  \vfill
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \setlength{\leftmargini}{0pt}
      \begin{itemize}
      \item<3-> \alert{Stratum 1}: predicates with no negation
      \item<3-> \alert{Stratum 2}: new predicates s.t. negative literals refer to
        predicates in \alert{Stratum~1}
      \end{itemize}
      \vspace{5em}
    \end{column}
    \begin{column}{0.6\textwidth}
      \hspace{-6.92em}
      \begin{tikzpicture}
        \node[draw,ellipse] (friend) {\textcolor{predicate}{friend}};
        \node[draw,ellipse,right=of friend] (influences) {\textcolor{predicate}{influences}};
        \node[draw,ellipse,below=of influences] (stress) {\textcolor{predicate}{stress}};
        \node[draw,ellipse,below=of stress] (spont) {\textcolor{predicate}{cancer\_spont}};
        \node[draw,ellipse,below=of spont] (smoke) {\textcolor{predicate}{cancer\_smoke}};
        \node[draw,ellipse,left=of spont] (person) {\textcolor{predicate}{person}};
        \node[draw,ellipse,right=of spont] (cancer) {\textcolor{predicate}{cancer}};
        \node[draw,ellipse,above=of cancer] (smokes) {\textcolor{predicate}{smokes}};
        \draw[-{Latex[length=2mm]}] (friend) -- (influences);
        \draw[-{Latex[length=2mm]}] (influences) -- (smokes);
        \draw[-{Latex[length=2mm]}] (smokes) -- (cancer);
        \draw[-{Latex[length=2mm]}] (stress) -- (smokes);
        \draw[-{Latex[length=2mm]}] (spont) -- (cancer);
        \draw[-{Latex[length=2mm]}] (smoke) -- (cancer);
        \draw[-{Latex[length=2mm]}] (person) -- (stress);
        \draw[-{Latex[length=2mm]}] (person) -- (spont);
        \draw[-{Latex[length=2mm]}] (person) -- (smoke);
        \draw<2->[-{Latex[length=2mm]},draw=red] (smokes) -- (friend);
      \end{tikzpicture}
    \end{column}
  \end{columns}
\end{frame}

@@

\section{Probabilistic Inference}

\begin{frame}{Inference Algorithms and Knowledge Compilation Maps}
  \begin{description}
  \item[NNF] negation normal form
  \item[BDD] binary decision diagrams
  \item[SDD] sentential decision diagrams
  \item[$k$-Best] only use the \alert{$k$} most probable proofs
  \item[d-DNNF] deterministic decomposable negation normal form
    \begin{itemize}
    \item<2-> for every pair \alert{$\alpha \lor \beta$}, we have \alert{$\alpha
        \land \beta = \bot$}
    \item<2-> for every pair \alert{$\alpha \land \beta$}, no atoms are shared
      between \alert{$\alpha$} and \alert{$\beta$}
    \end{itemize}
  \end{description}
\end{frame}

\begin{frame}{Example Diagrams for \alert{$C \land (A \lor \neg B)$}}
  \begin{columns}[b]
    \begin{column}{0.33\textwidth}
      \begin{figure}
        \centering
        \begin{tikzpicture}[level distance=0.8cm]
          \node[draw,circle] {$\land$}
          child {node[draw,rectangle] {$C$}}
          child {node[draw,circle] {$\lor$}
            child {node[draw,rectangle] {$A$}}
            child {node[draw,rectangle] {$\neg B$}
            }
          };
        \end{tikzpicture}
        \caption{NNF}
      \end{figure}
      \vspace{-1em}
      \begin{figure}
      \centering
        \begin{tikzpicture}[level distance=0.8cm]
          \node[draw,circle] (A) {$A$}
          child {edge from parent[draw=none]}
          child {node[draw,circle] (B) {$B$} edge from parent[dashed]
            child {node[draw,circle,solid] (C) {$C$} edge from parent[dashed]
              child {node[draw,rectangle,solid] {$1$} edge from parent[solid]}
              child {node[draw,rectangle,solid] (0) {$0$}}
            }
            child {edge from parent[draw=none]}
          };
          \draw (A) -- (C);
          \draw (B) -- (0);
        \end{tikzpicture}
        \caption{BDD}
      \end{figure}
    \end{column}
    \begin{column}{0.33\textwidth}
      \begin{figure}
        \centering
        \begin{tikzpicture}[level distance=0.8cm]
          \node[draw,circle] {$\land$}
          child {node[draw,circle] {$\lor$}
            child {node[draw,circle] (parent) {$\land$}
              child {node[draw,rectangle] {$B$}}
              child {edge from parent[draw=none]}
            }
            child {node[draw,circle] {$\land$}
              child {node[draw,circle] {$\lor$}
                child {node[draw,rectangle] (A) {$A$}}
                child {node[draw,rectangle] {$\neg A$}}
              }
              child {node[draw,rectangle] {$\neg B$}}
            }
          }
          child {node[draw,rectangle] {$C$}};
          \draw (parent) -- (A);
        \end{tikzpicture}
        \caption{d-DNNF}
      \end{figure}
    \end{column}
    \begin{column}{0.33\textwidth}
      \begin{figure}
        \centering
        \begin{tikzpicture}[level distance=0.8cm]
          \tikzset{
            mysplit/.style={
              draw,
              rectangle,
              rectangle split,
              rectangle split horizontal,
              rectangle split parts=2
            }
          }
          \node[draw,circle] {$1$}
          child {node[mysplit] (bullet) {
              \nodepart{one} $\neg A$
              \nodepart{two}
            }
            child {node[draw,circle] (3) {$3$} edge from parent[draw=none]
              child {node[mysplit] {
                  \nodepart{one} $\neg B$
                  \nodepart{two} $C$
                }
              }
              child {node[mysplit] {
                  \nodepart{one} $B$
                  \nodepart{two} $\bot$
                }
              }
            }
          }
          child {node[mysplit] {
              \nodepart{one} $A$
              \nodepart{two} $C$
            }};
          \draw[*-] let \p1 = (bullet.two), \p2 = (bullet.center) in ({\x1 + 2.5},{\y2 + 2}) -- (3);
        \end{tikzpicture}
        \caption{SDD}
      \end{figure}
      \vspace{-2em}
      \begin{figure}
      \centering
      \begin{tikzpicture}[level distance=0.8cm]
        \node[draw,circle] {$1$}
        child {node[draw,circle] {$0$}
          child {node[draw,rectangle] {$A$}}
        }
        child {node[draw,circle] {$3$}
          child {node[draw,circle] {$2$}
            child {node[draw,rectangle] {$B$}}
          }
          child {node[draw,circle] {$4$}
            child {node[draw,rectangle] {$C$}}
          }
        };
      \end{tikzpicture}
      \caption{vtree}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

@@

\section{Summary}

\begin{frame}{Summary}
  \begin{itemize}
  \item The model can generate (approximately) realistic instances of reasonable
    size
  \item The main performance bottleneck can be addressed by generating programs
    with a simpler structure
  \item Open questions and future work
    \begin{itemize}
    \item Can the model be used to ensure uniform sampling?
    \item What is the reason behind all algorithms behaving similarly?
    \item Why does independence have no effect on inference time?
    \item Can random program generation be useful in, e.g., learning?
    \end{itemize}
  \end{itemize}
  \begin{block}{The implementation of the model is available at}
    \url{https://github.com/dilkas/random-logic-programs}
  \end{block}
\end{frame}
